---
title: "Fires in the Paraná River Delta, Argentina"
author: "Natalia Morandeira (nmorandeira@unsam.edu.ar)"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=F}
knitr::opts_chunk$set(echo = F, warning = F)
```

--------
## Data source

```{r set_ZIP_names, include=F}
zipfiles <- list.files(path = "data/zip/")
zipfiles
```


This report is generated from the following zip files: __`r zipfiles`__ and a polygon of the study area.

The naming conventions of the FIRMS data are: 

  - DL_FIRE_M6.xx if you requested MODIS data (M6 stands for MODIS Collection 6), or 
  - DL_FIRE_V1.xx if you requested VIIRS 375m data from S-NPP.
  
Each of these zips has two shapefiles: recent and archive records.

```{r unzip, eval=T, include = F}
MODISzip <- paste("data/zip/", zipfiles[1], sep="")
VIIRSzip <- paste("data/zip/", zipfiles[2], sep="")

unzip(MODISzip, exdir = "data/hotspots/") 
unzip(VIIRSzip, exdir = "data/hotspots/") 
```

```{r read_spatial_data, eval=T, include = F}
library(sf)
library(stringr)

study_area <- list.files(path = "data/study_area/")
study_area_shp <- str_detect(study_area, "shp")=="TRUE"
study_area <- subset(study_area, study_area_shp == "TRUE")
study_area <- paste("data/study_area/", study_area, sep="")
study_area <- st_read(study_area, quiet=TRUE)

spatial_files <- list.files(path = "data/hotspots")
shp_extension <- str_detect(spatial_files, "shp")=="TRUE"
spatial_files <- subset(spatial_files, shp_extension == "TRUE")

hotspots_VIIRS_recent <- which(str_detect(spatial_files, "fire_nrt_V1")=="TRUE")
hotspots_VIIRS_recent <- paste("data/hotspots/", spatial_files[hotspots_VIIRS_recent], sep="")
hotspots_VIIRS_recent <- st_read(hotspots_VIIRS_recent, quiet=TRUE)

hotspots_VIIRS_archive <- which(str_detect(spatial_files, "fire_archive_V1")=="TRUE")
hotspots_VIIRS_archive <- paste("data/hotspots/", spatial_files[hotspots_VIIRS_archive], sep="")
hotspots_VIIRS_archive <- st_read(hotspots_VIIRS_archive, quiet=TRUE)

hotspots_MODIS_recent <- which(str_detect(spatial_files, "fire_nrt_M6")=="TRUE")
hotspots_MODIS_recent <- paste("data/hotspots/", spatial_files[hotspots_MODIS_recent], sep="")
hotspots_MODIS_recent <- st_read(hotspots_MODIS_recent, quiet=TRUE)

hotspots_MODIS_archive <- which(str_detect(spatial_files, "fire_archive_M6")=="TRUE")
hotspots_MODIS_archive <- paste("data/hotspots/", spatial_files[hotspots_MODIS_archive], sep="")
hotspots_MODIS_archive <- st_read(hotspots_MODIS_archive, quiet=TRUE)


```




```{r merge_shp, eval=T, include = F}
#Remove CONFIDENCE columns since they have different formats
hotspots_VIIRS_recent$CONFIDENCE = NULL
hotspots_VIIRS_archive$CONFIDENCE = NULL
hotspots_VIIRS_archive$TYPE = NULL

hotspots_MODIS_recent$CONFIDENCE = NULL
hotspots_MODIS_archive$CONFIDENCE = NULL
hotspots_MODIS_archive$TYPE = NULL

#Merge
hotspots_v <- rbind(hotspots_VIIRS_recent,hotspots_VIIRS_archive)
hotspots_m <- rbind(hotspots_MODIS_recent, hotspots_MODIS_archive)
colnames(hotspots_v) <- colnames(hotspots_m)
hotspots_all <- rbind(hotspots_v, hotspots_m)

```


```{r reproject,  eval=T, include = F}
library(sf)
study_area_crs <- st_crs(study_area)
#study_area_epsg <- study_area_crs$epsg #check, is not working
study_area_epsg <- 5347
study_area <- st_transform(study_area, study_area_epsg)

#reproject
hotspots_all <- st_transform(hotspots_all, study_area_epsg)
```

```{r clip, eval=T, include = F}
hotspots_studyarea <- st_intersects(x=hotspots_all, y=study_area)
hotspots_studyarea_logical <- lengths(hotspots_studyarea) > 0

hotspots_all = hotspots_all[hotspots_studyarea_logical, ]

```

```{r save_shp, eval=F, include =F}
sf::st_write(hotspots_all, "output/hotspot_all_2020.gpkg")

```


```{r filter_VIIRS_2020,  eval=T, include = F}

library(tidyverse)
library(spdplyr) # required to run dplyr funcions on spatial objects

hotspots_VIIRS_2020 <- hotspots_all %>%
  mutate(year = format(ACQ_DATE, "%Y")) %>%
  filter(year == "2020") %>%
  filter(INSTRUMENT == "VIIRS")

last_date <- max(hotspots_VIIRS_2020$ACQ_DATE)

hotspots_count2020 <- as.numeric(nrow(hotspots_VIIRS_2020)) # Total hotspots in 2020
```

## Summary results and plots

The number of VIIRS hotspots recorded during this year is **`r format(hotspots_count2020, scientific=F)`**, up to **`r last_date`**.

```{r save_shp2, eval=F, include =F}
sf::st_write(hotspots_VIIRS_2020, "output/hotspot_VIIRS_2020.gpkg")

```

This interactive map shows the location of the hotspot records. The date of the record can be obtained by positioning the mouse onto the point.

```{r map}
library(tmap)
tmap_mode("view") 
mapa <- tm_shape(hotspots_VIIRS_2020,  bbox= study_area) + 
  tm_dots(col = "red", size = 0.03, alpha = 0.3, id="ACQ_DATE") +
  tm_layout(title = "Focos de incendio en el Bajo Paraná - 2020, Datos VIIRS", legend.frame = T) +
  tm_shape(study_area) +
  tm_borders(col = "grey60", lwd = 2) 

mapa
```



```{r cum, eval=T, include = F}
library(janitor)
focos2020 <- hotspots_VIIRS_2020
st_geometry(focos2020) <- NULL #quitar geometría
focos2020 <- clean_names(focos2020)
colnames(focos2020)

focos_cum <- focos2020 %>%
  mutate(cantidad = 1) %>%
  group_by(acq_date) %>%
  summarize(cantidad_diaria = sum(cantidad)) 
focos_cum$acumulado <- cumsum(focos_cum$cantidad_diaria)
head(focos_cum)
```

First, we'll get a plot of the daily records. Spanish and English versions

```{r daily}
#Spanish
plot_diario <- focos_cum %>%
  pivot_longer(names_to = "focos", values_to = "cantidad", col=cantidad_diaria:acumulado) %>%
  filter(focos == "cantidad_diaria") %>%
  ggplot(aes (x=acq_date, y=cantidad, color=focos)) +
  geom_col() +
  scale_color_manual(values=c("navyblue", "navyblue"),  labels = c("Focos diarios", "Nuevos focos")) +
  xlab("Fecha") +
  ylab("Cantidad de focos de calor") +
  scale_x_date(date_labels = "%d/%m/%Y", breaks = "week") +
  scale_y_continuous(breaks=seq(from = 0, to = 3000, by = 200)) +
  theme_bw()+
  theme(axis.text.x = element_text(angle=45, hjust=1, color="black"), plot.caption = element_text(hjust = 0, vjust=1, face = "italic"), plot.title = element_text(face = "bold"), axis.line=element_line(color="black"), axis.text.y=element_text(color="black"), legend.position = "none" ) +
  labs(title="Incendios en el Delta del Paraná (Argentina) durante 2020", subtitle = "Focos de calor por día, en base a datos VIIRS de FIRMS-NASA", caption = "Natalia Morandeira; CONICET & 3iA-UNSAM")  

plot_diario

plot_daily_eng <- focos_cum %>%
  pivot_longer(names_to = "focos", values_to = "cantidad", col=cantidad_diaria:acumulado) %>%
  filter(focos == "cantidad_diaria") %>%
  ggplot(aes (x=acq_date, y=cantidad, color=focos)) +
  geom_col() +
  scale_color_manual(values=c("navyblue", "navyblue"),  labels = c("New hotspots", "New hotspots")) +
  xlab("Date") +
  ylab("Number of VIIRS hotspots") +
  scale_x_date(date_labels = "%m/%d/%Y", breaks = "week") +
  scale_y_continuous(breaks=seq(from = 0, to = 3000, by = 200)) +
  theme_bw()+
  theme(axis.text.x = element_text(angle=45, hjust=1, color="black"), plot.caption = element_text(hjust = 0, vjust=1, face = "italic"), plot.title = element_text(face = "bold"), legend.position="none", axis.line=element_line(color="black"), axis.text.y=element_text(color="black")) +
  labs(title="Potential fires at the Paraná River Delta (Argentina)", subtitle = "Daily hotspots, based on VIIRS data - FIRMS-NASA", caption = "Natalia Morandeira; CONICET & 3iA-UNSAM")  

plot_daily_eng
```



```{r save_daily, eval=T, include=F}
ggsave("output/Focos_diario_2020-12-31.png", plot = plot_diario, width = 9, height = 5, dpi = 300)
ggsave("output/Hotspot_daily_2020-12-31.png", plot = plot_daily_eng, width = 9, height = 5, dpi = 300)
```

A plot of the cumulative hostspots. Spanish and English versions.

```{r plot_cum}
#Spanish
plot_acum <- focos_cum %>%
  pivot_longer(names_to = "focos", values_to = "cantidad", col=cantidad_diaria:acumulado) %>%
  ggplot(aes (x=acq_date, y=cantidad, color=focos)) +
  geom_line() +
  scale_color_manual(values=c("darkred", "navyblue"),  labels = c("Focos acumulados", "Focos activos")) +
  xlab("Fecha") +
  ylab("Cantidad de focos de calor") +
  scale_x_date(date_labels = "%d/%m/%Y", breaks = "week") +
  scale_y_continuous(breaks=seq(from = 0, to = (hotspots_count2020 + 1000), by = 2500)) +
  theme_bw()+
  theme(axis.text.x = element_text(angle=45, hjust=1, color="black"), plot.caption = element_text(hjust = 0, vjust=1, face = "italic"), plot.title = element_text(face = "bold"), legend.position=c(0.12, 0.85), legend.title = element_blank(), axis.line=element_line(color="black"), axis.text.y=element_text(color="black") , legend.background = element_rect(colour ="grey40", size = 0.2)) +
  labs(title="Incendios en el Delta del Paraná (Argentina) durante 2020", subtitle = "Focos de calor por día y acumulados al 31/12/2020, en base a datos VIIRS de FIRMS-NASA", caption = "Natalia Morandeira; CONICET & 3iA-UNSAM")  

plot_acum


#idem en inglés
plot_cum_eng <- focos_cum %>%
  pivot_longer(names_to = "focos", values_to = "cantidad", col=cantidad_diaria:acumulado) %>%
  ggplot(aes (x=acq_date, y=cantidad, color=focos)) +
  #annotation_custom(rasterGrob(img, width = unit(1,"npc"), height = unit(1,"npc")), 
  #                  -Inf, Inf, -Inf, Inf) +
  geom_line() +
  scale_color_manual(values=c("darkred", "navyblue"),  labels = c("Cumulative hotspots", "New hotspots")) +
  xlab("Date") +
  ylab("Number of VIIRS hotspots") +
  scale_x_date(date_labels = "%m/%d/%Y", breaks = "week") +
  scale_y_continuous(breaks=seq(from = 0, to = (hotspots_count2020 + 1000), by = 2500)) +
  theme_bw()+
  theme(axis.text.x = element_text(angle=45, hjust=1, color="black"), plot.caption = element_text(hjust = 0, vjust=1, face = "italic"), plot.title = element_text(face = "bold"), legend.position=c(0.15, 0.85), legend.title = element_blank(), axis.line=element_line(color="black"), axis.text.y=element_text(color="black") , legend.background = element_rect(colour ="grey40", size = 0.2)) +
  labs(title="Potential fires at the Paraná River Delta (Argentina)", subtitle = "Daily hotspots and cumulative numbers up to 12/31/2020, based on VIIRS data - FIRMS-NASA", caption = "Natalia Morandeira; CONICET & 3iA-UNSAM")  
plot_cum_eng

```


```{r save_cum, eval=T, include=F}
ggsave("output/Focos_acumulados_2020-12-31.png", plot = plot_acum, width = 9, height = 5, dpi = 300)
ggsave("output/Hotspot_cum_2020-12-31.png", plot = plot_cum_eng, width = 9, height = 5, dpi = 300)
```

The next plot summarizes the number of VIIRS hotspots per month. 

```{r VIIRS_mes}
focos_VIIRS <- hotspots_all
focos_VIIRS <- clean_names(focos_VIIRS)

focos_VIIRS <- focos_VIIRS %>%
  filter(instrument == "VIIRS") %>%
  mutate(cantidad = 1) %>%
  group_by(acq_date) %>%
  summarize(cantidad_diaria = sum(cantidad)) 

focos_VIIRS2020_mes <- focos_VIIRS %>% 
  filter (acq_date > "2020-01-01") %>%
  mutate(mes = format(acq_date, "%Y/%m")) %>%
  group_by(mes) %>%
  summarize(cantidad_mes = sum(cantidad_diaria))

#convertir a fecha
focos_VIIRS2020_mes$mes <- as.Date(paste(focos_VIIRS2020_mes$mes,1,sep="/"),"%Y/%m/%d") 
#head(focos_VIIRS2020_mes) 

VIIRS2020_mes <- ggplot(focos_VIIRS2020_mes, aes(x=mes, y=cantidad_mes)) +
  geom_col( fill="lightblue") +
  geom_text(aes(label = cantidad_mes), col="black") +
  xlab("mes") +
  ylab("Cantidad de focos VIIRS por mes") +
  theme_bw()
VIIRS2020_mes

VIIRSagosto <- (focos_VIIRS2020_mes$cantidad_mes)[8]

ggsave("output/Focos_VIIRS2020_mensual.png", plot = VIIRS2020_mes, width = 8, height = 5, dpi = 300)
```

The month with the highest number of hotspots is August, with **`r format(VIIRSagosto, scientific=F)`** hotspots (**`r round(VIIRSagosto/hotspots_count2020*100,1)`%** % of the total hotspots up to **`r last_date`**).


## Historical comparison using MODIS (November 2001 - present)
While VIIRS data (resolution: 375 m) is available from 2012, MODIS data is available from 2001. However, MODIS resolution is 1 km, so fewer hotspots are reported and each hotspot corresponds to a greater area. 

```{r MODIS}
focos_MODIS <- hotspots_all

focos_MODIS <- clean_names(focos_MODIS)
#colnames(focos_MODIS)

focos_MODIS <- focos_MODIS %>%
  filter(instrument == "MODIS") %>%
  mutate(cantidad = 1) %>%
  group_by(acq_date) %>%
  summarize(cantidad_diaria = sum(cantidad)) 

#head(focos_MODIS)
```

Now we compute and plot the number of MODIS hotspots per year.

```{r modis2}
#calculo de focos por anio
focos_MODIS_anio <- focos_MODIS %>% 
  mutate(anio = format(acq_date, "%Y")) %>%
  group_by(anio) %>%
  summarize(cantidad_anio = sum(cantidad_diaria))

#glimpse(focos_MODIS_anio)

MODIS_year <- ggplot(subset(focos_MODIS_anio, focos_MODIS_anio$anio!="2000"), aes(x=anio, y=cantidad_anio)) +
  geom_col( fill="lightblue") +
  geom_text(aes(label = cantidad_anio), col="black") +
  xlab("Año") +
  ylab("Cantidad de focos MODIS por año") +
  theme_bw()
MODIS_year

ggsave("output/MODIS_2001-2020.png", plot = MODIS_year, width = 9, height = 5, dpi = 300)

MODIS_2020 <- as.numeric(as.list(subset(focos_MODIS_anio, focos_MODIS_anio$anio ==2020))[2])
MODIS_2008 <-  as.numeric(as.list(subset(focos_MODIS_anio, focos_MODIS_anio$anio ==2008))[2])
```

The cumulative number of MODIS hotspots during 2020 is **`r MODIS_2020`** up to **`r last_date`**, which is **`r round(MODIS_2020/MODIS_2008*100,1)`%** of the MODIS hotspots recorded during 2008. 

Some plots summarizing the historical hotspots from MODIS data. Daily and monthly plots.

```{r MODIS_plots}
#diario
MODIS_diario <- focos_MODIS %>%
  ggplot(aes (x=acq_date, y=cantidad_diaria)) +
  geom_line(color="darkred") +
  xlab("Fecha") +
  ylab("Cantidad de focos de incendio") +
  scale_x_date(date_labels = "%d/%m/%Y", breaks = "year") +
  theme(axis.text.x = element_text(angle=45, hjust=1, color="black"), plot.caption = element_text(hjust = 0, vjust=1, face = "italic"), plot.title = element_text(face = "bold"), legend.position=c(0.12, 0.85), legend.title = element_blank(), axis.line=element_line(color="black"), axis.text.y=element_text(color="black")) +
  labs(title="Datos históricos - incendios en el Delta del Paraná (Argentina)", subtitle = "Focos de calor por día (2001 - presente), en base a datos MODIS de FIRMS-NASA", caption = "Natalia Morandeira; 3iA-UNSAM") 

MODIS_diario

#ggsave("Focos_historicos_diarios.png", plot = p3, width = 8, height = 5, dpi = 300)
```

```{r MODIS_month}
#agrupar dias en meses
focos_MODIS_mes <- focos_MODIS %>% 
  mutate(mes = format(acq_date, "%Y/%m")) %>%
  group_by(mes) %>%
  summarize(cantidad_mes = sum(cantidad_diaria))

#convertir a fecha
focos_MODIS_mes$mes <- as.Date(paste(focos_MODIS_mes$mes,1,sep="/"),"%Y/%m/%d")
head(focos_MODIS_mes)

MODIS_mes <-  focos_MODIS_mes %>%
  ggplot(aes (x=mes, y=cantidad_mes)) +
  geom_line(col="darkred") +
  scale_color_manual(values=c("red", "blue"),  labels = c("Focos acumulados", "Nuevos focos")) +
  xlab("Fecha") +
  ylab("Cantidad de focos de incendio") +
  scale_x_date(date_labels = "%m/%Y", breaks = "6 months") +
  theme(axis.text.x = element_text(angle=45, hjust=1, color="black"), plot.caption = element_text(hjust = 0, vjust=1, face = "italic"), plot.title = element_text(face = "bold"), legend.position=c(0.12, 0.85), legend.title = element_blank(), axis.line=element_line(color="black"), axis.text.y=element_text(color="black")) +
  labs(title="Datos históricos - incendios en el Delta del Paraná (Argentina)", subtitle = "Focos de calor mensuales (2001 - presente), en base a datos MODIS de FIRMS-NASA", caption = "Natalia Morandeira; 3iA-UNSAM") 

MODIS_mes

#ggsave("Focos_historicos_mensual.png", plot = p4, width = 8, height = 5, dpi = 300)
```

Now we can compare VIIRS and MODIS records, from 2012.

```{r VIIRS_all}
focos_VIIRS <- hotspots_all
focos_VIIRS <- clean_names(focos_VIIRS)

focos_VIIRS <- focos_VIIRS %>%
  filter(instrument == "VIIRS") %>%
  mutate(cantidad = 1) %>%
  group_by(acq_date) %>%
  summarize(cantidad_diaria = sum(cantidad)) 

#calculo de focos por anio
focos_VIIRS_anio <- focos_VIIRS %>% 
  mutate(anio = format(acq_date, "%Y")) %>%
  group_by(anio) %>%
  summarize(cantidad_anio = sum(cantidad_diaria))

#glimpse(focos_VIIRS_anio)
#write.csv(x = focos_VIIRS_anio, file = "focos_anuales_VIIRS.csv")


VIIRS_anio <- ggplot(subset(focos_VIIRS_anio, focos_VIIRS_anio$anio!="2000"), aes(x=anio, y=cantidad_anio)) +
  geom_col( fill="lightblue") +
  geom_text(aes(label = cantidad_anio), col="black") +
  xlab("Año") +
  ylab("Cantidad de focos VIIRS por año") +
  theme_bw()
VIIRS_anio

ggsave("output/VIIRS_2012-2020.png", plot = VIIRS_anio, width = 9, height = 5, dpi = 300)
```

Lastly, we can comparte VIIRS and MODIS records. Spanish and English versions. 

```{r comparison}
#preparo label del eje X
break.fecha <- c(seq(from = as.Date("2001-01-01"), to = as.Date("2020-11-01"), by = "6 months"), as.Date("2021-01-01"))

#agrupar dias en meses
focos_VIIRS_mes <- focos_VIIRS %>% 
  mutate(mes = format(acq_date, "%Y/%m")) %>%
  group_by(mes) %>%
  summarize(cantidad_mes = sum(cantidad_diaria))

#convertir a fecha
focos_VIIRS_mes$mes <- as.Date(paste(focos_VIIRS_mes$mes,1,sep="/"),"%Y/%m/%d")
#head(focos_VIIRS_mes)

MODIS_VIIRS_plot <-  ggplot(data = focos_MODIS_mes, aes (x=mes, y=cantidad_mes)) +
  geom_line(col="darkred") +
  xlab("Fecha") +
  ylab("Cantidad de focos de incendio") +
  scale_x_date(date_labels = "%m/%Y", breaks = break.fecha) +
  theme_bw() +
  theme(axis.text.x = element_text(angle=45, hjust=1, color="black"), plot.caption = element_text(hjust = 0, vjust=1, face = "italic"), plot.title = element_text(face = "bold"), legend.position=c(0.12, 0.85), legend.title = element_blank(), axis.line=element_line(color="black"), axis.text.y=element_text(color="black")) +
  labs(title="Datos históricos - incendios en el Delta del Paraná (Argentina)", subtitle = "Focos de calor mensuales (2001 - presente), en base a datos FIRMS-NASA. Las líneas indican \nfocos detectados por MODIS (1 km) y las barras representan focos VIIRS (375 m)", caption = "Natalia Morandeira; 3iA-UNSAM") 

MODIS_VIIRS_plot<- MODIS_VIIRS_plot + geom_col(data= focos_VIIRS_mes, aes (x=mes, y=cantidad_mes, group=1),  fill="orange", alpha=0.5, width = 30)

MODIS_VIIRS_plot



MODIS_VIIRS_plot_eng <-  ggplot(data = focos_MODIS_mes, aes (x=mes, y=cantidad_mes)) +
  geom_line(col="darkred") +
  xlab("Date") +
  ylab("Number of hotspots") +
  scale_x_date(date_labels = "%m/%Y", breaks = break.fecha) +
  theme_bw() +
  theme(axis.text.x = element_text(angle=45, hjust=1, color="black"), plot.caption = element_text(hjust = 0, vjust=1, face = "italic"), plot.title = element_text(face = "bold"), legend.position=c(0.12, 0.85), legend.title = element_blank(), axis.line=element_line(color="black"), axis.text.y=element_text(color="black")) +
  labs(title="Historical data - Potential fires in the Paraná River Delta (Argentina)", subtitle = "Monthly number of hotspots (2001 - present), based on FIRMS-NASA data. \nLines indicate MODIS hotspots (1 km) and bars indicate VIIRS hotspots (375 m)", caption = "Patricia Kandus, Natalia Morandeira and Priscilla Minotti; 3iA-UNSAM") 

MODIS_VIIRS_plot_eng <- MODIS_VIIRS_plot + geom_col(data= focos_VIIRS_mes, aes (x=mes, y=cantidad_mes, group=1),  fill="orange", alpha=0.5, width = 30)

MODIS_VIIRS_plot_eng


```

The plots are saved to the output folder.