---
title: "Fires in the Paraná River Delta"
author: "Natalia Morandeira"
date: "8/7/2020"
output: html_document
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE)
```

--------
## Data source
- Paraná River Delta limits: shapefile of the study area
- Hotspots from FIRMS-NASA, https://firms.modaps.eosdis.nasa.gov/map/#t:adv;d:2020-07-28..2020-08-03;@-59.9,-33.1,8z

### Naming conventions of the FIRMS data: 

  - DL_FIRE_M6.xx if you requested MODIS data (M6 stands for MODIS Collection 6), or 
  - DL_FIRE_V1.xx if you requested VIIRS 375m data from S-NPP
  - DL_FIRE_J1.xx if you requested VIIRS 375m data from NOAA-20(JPSS-1)

### Spatial data pre-processing

1. Read files and unzip
1. Handle the files as spatial data
1. Merge
1. Re-project
1. Clip by study area 

To read and unzip, we first set the names of the two ZIP files, one from MODIS and the other from VIIRS satellites. Each of these zips has two shapefiles: recent and archive records.

```{r set_ZIP_names, eval=T, include=T}
zipfiles <- list.files(path = "data/zip/")
length(zipfiles)
```

```{r unzip, eval=F, include = T}
MODISzip <- paste("data/zip/", files[1], sep="")
VIIRSzip <- paste("data/zip/", files[2], sep="")

unzip(MODISzip, exdir = "data/hotspots/") 
unzip(VIIRSzip, exdir = "data/hotspots/") 
```

There are `r length(zipfiles)` zip files in the database (expected: two files).

Handle spatial data. Four shapefiles + Study area.
The code is prepared to detect the appropriate strings and generate the spatial objects.

```{r read_spatial_data, eval=F, include = T}
library(sf)
library(stringr)
spatial_files <- list.files(path = "data/hotspots")
shp_extension <- str_detect(spatial_files, "shp")=="TRUE"
spatial_files <- subset(spatial_files, shp_extension == "TRUE")
  

hostpots_VIIRS_recent <- which(str_detect(spatial_files, "fire_nrt_V1")=="TRUE")
hostpots_VIIRS_recent <- paste("data/hotspots/", spatial_files[hostpots_VIIRS_recent], sep="")
hostpots_VIIRS_recent <- st_read(hostpots_VIIRS_recent, quiet=TRUE)

hostpots_VIIRS_archive <- which(str_detect(spatial_files, "fire_archive_V1")=="TRUE")
hostpots_VIIRS_archive <- paste("data/hotspots/", spatial_files[hostpots_VIIRS_archive], sep="")
hostpots_VIIRS_archive <- st_read(hostpots_VIIRS_archive, quiet=TRUE)


hostpots_MODIS_recent <- which(str_detect(spatial_files, "fire_nrt_M6")=="TRUE")
hostpots_MODIS_recent <- paste("data/hotspots/", spatial_files[hostpots_MODIS_recent], sep="")
hostpots_MODIS_recent <- st_read(hostpots_MODIS_recent, quiet=TRUE)


hostpots_MODIS_archive <- which(str_detect(spatial_files, "fire_archive_M6")=="TRUE")
hostpots_MODIS_archive <- paste("data/hotspots/", spatial_files[hostpots_MODIS_archive], sep="")
hostpots_MODIS_archive <- st_read(hostpots_MODIS_archive, quiet=TRUE)


study_area <- st_read("data/Delta_BajoParana_EPSG5347.shp", quiet=TRUE)
   
```

**Work in progress** Now we need to merge the shapefiles, reproject them to EPSG 5347 and clip with the study area shapefile.

```{r merge_shp, eval=F}
#Remove CONFIDENCE columns since they have different formats
hostpots_VIIRS_recent$CONFIDENCE = NULL
hostpots_VIIRS_archive$CONFIDENCE = NULL
hostpots_VIIRS_archive$TYPE = NULL

hostpots_MODIS_recent$CONFIDENCE = NULL
hostpots_MODIS_archive$CONFIDENCE = NULL
hostpots_MODIS_archive$TYPE = NULL

```

### Importing a pre-processed hotspot layer as spatial object

Alternatively, the script can also begin with a pre-processed final hotspot layer.

```{r read_file, eval = T, include = TRUE}
library(sf)
hotspots_all <- st_read("data/hotspots/focos_all_20201110.shp", quiet=TRUE) 
study_area <- st_read(dsn = "data/Delta_BajoParana_EPSG5347.shp", quiet=TRUE)

```


### Data processing

This layer has both MODIS and VIIRS hotspots.
Extract VIIRS 2020 data.

```{r filter_VIIRS_2020}

library(tidyverse)
library(spdplyr) # required to run dplyr funcions on spatial objects

hotspots_VIIRS_2020 <- hotspots_all %>%
  mutate(year = format(ACQ_DATE, "%Y")) %>%
  filter(year == "2020") %>%
  filter(INSTRUMENT == "VIIRS")

last_date <- max(hotspots_VIIRS_2020$ACQ_DATE)

hotspots_count2020 <- as.numeric(nrow(hotspots_VIIRS_2020)) # Total hotspots in 2020
```

The number of VIIRS hotspots recorded during this year is **`r hotspots_count2020`**, up to **`r last_date`**.

Let's get a quick interactive map to visualize the location of the hotspot records. The date of the record can be obtained by positioning the mouse onto the point.

```{r map}
library(tmap)
tmap_mode("view") 
mapa <- tm_shape(hotspots_VIIRS_2020,  bbox= study_area) + 
  tm_dots(col = "red", alpha = 0.3, id="ACQ_DATE") +
  tm_layout(title = "Focos de incendio en el Bajo Paraná - 2020, Datos VIIRS", legend.frame = T) +
  tm_shape(study_area) +
  tm_borders(col = "grey60", lwd = 2) 

mapa
```


We need to generate cummulative number of hotspot records during 2020.

```{r cum}
library(janitor)
focos2020 <- hotspots_VIIRS_2020
st_geometry(focos2020) <- NULL #quitar geometría
focos2020 <- clean_names(focos2020)
colnames(focos2020)

focos_cum <- focos2020 %>%
  mutate(cantidad = 1) %>%
  group_by(acq_date) %>%
  summarize(cantidad_diaria = sum(cantidad)) 
focos_cum$acumulado <- cumsum(focos_cum$cantidad_diaria)
head(focos_cum)
```

First, we'll get a plot of the daily records. Spanish and English versions

```{r daily}
#Spanish
plot_diario <- focos_cum %>%
  pivot_longer(names_to = "focos", values_to = "cantidad", col=cantidad_diaria:acumulado) %>%
  filter(focos == "cantidad_diaria") %>%
  ggplot(aes (x=acq_date, y=cantidad, color=focos)) +
  geom_col() +
  scale_color_manual(values=c("navyblue", "navyblue"),  labels = c("Focos diarios", "Nuevos focos")) +
  xlab("Fecha") +
  ylab("Cantidad de focos de calor") +
  scale_x_date(date_labels = "%d/%m/%Y", breaks = "week") +
  scale_y_continuous(breaks=seq(from = 0, to = 3000, by = 200)) +
  theme_bw()+
  theme(axis.text.x = element_text(angle=45, hjust=1, color="black"), plot.caption = element_text(hjust = 0, vjust=1, face = "italic"), plot.title = element_text(face = "bold"), axis.line=element_line(color="black"), axis.text.y=element_text(color="black"), legend.position = "none" ) +
  labs(title="Incendios en el Delta del Paraná (Argentina) durante 2020", subtitle = "Focos de calor por día, en base a datos VIIRS de FIRMS-NASA", caption = "Natalia Morandeira; 3iA-UNSAM")  

plot_diario

plot_daily_eng <- focos_cum %>%
  pivot_longer(names_to = "focos", values_to = "cantidad", col=cantidad_diaria:acumulado) %>%
  filter(focos == "cantidad_diaria") %>%
  ggplot(aes (x=acq_date, y=cantidad, color=focos)) +
  geom_col() +
  scale_color_manual(values=c("navyblue", "navyblue"),  labels = c("New hotspots", "New hotspots")) +
  xlab("Date") +
  ylab("Number of VIIRS hotspots") +
  scale_x_date(date_labels = "%m/%d/%Y", breaks = "week") +
  scale_y_continuous(breaks=seq(from = 0, to = 3000, by = 200)) +
  theme_bw()+
  theme(axis.text.x = element_text(angle=45, hjust=1, color="black"), plot.caption = element_text(hjust = 0, vjust=1, face = "italic"), plot.title = element_text(face = "bold"), legend.position="none", axis.line=element_line(color="black"), axis.text.y=element_text(color="black")) +
  labs(title="Potential fires at the Paraná River Delta (Argentina)", subtitle = "Daily hotspots, based on VIIRS data - FIRMS-NASA", caption = "Natalia Morandeira; 3iA-UNSAM")  

plot_daily_eng
```

We may export these plots.

```{r save_daily, eval=F, include=T}
<<<<<<< HEAD
ggsave("output/Focos_diario_2020-11-10.png", plot = plot_diario, width = 8, height = 5, dpi = 300)
ggsave("output/Hotspot_daily_2020-11-10.png", plot = plot_daily_eng, width = 8, height = 5, dpi = 300)
=======
ggsave("output/Focos_diario_2020-10-04.png", plot = plot_diario, width = 8, height = 5, dpi = 300)
ggsave("output/Hotspot_daily_2020-10-04.png", plot = plot_daily_eng, width = 8, height = 5, dpi = 300)
>>>>>>> a8a7c5ead622c4e9943587b8aabe666ba7da5c88
```

A plot of the cummulative hostspots. Spanish and English versions.

```{r plot_cum}
#Spanish
plot_acum <- focos_cum %>%
  pivot_longer(names_to = "focos", values_to = "cantidad", col=cantidad_diaria:acumulado) %>%
  ggplot(aes (x=acq_date, y=cantidad, color=focos)) +
  #annotation_custom(rasterGrob(img, width = unit(1,"npc"), height = unit(1,"npc")), 
  #                  -Inf, Inf, -Inf, Inf) +
  geom_line() +
  scale_color_manual(values=c("darkred", "navyblue"),  labels = c("Focos acumulados", "Focos activos")) +
  xlab("Fecha") +
  ylab("Cantidad de focos de calor") +
  scale_x_date(date_labels = "%d/%m/%Y", breaks = "week") +
  scale_y_continuous(breaks=seq(from = 0, to = (hotspots_count2020 + 1000), by = 2500)) +
  theme_bw()+
  theme(axis.text.x = element_text(angle=45, hjust=1, color="black"), plot.caption = element_text(hjust = 0, vjust=1, face = "italic"), plot.title = element_text(face = "bold"), legend.position=c(0.12, 0.85), legend.title = element_blank(), axis.line=element_line(color="black"), axis.text.y=element_text(color="black") , legend.background = element_rect(colour ="grey40", size = 0.2)) +
  labs(title="Incendios en el Delta del Paraná (Argentina) durante 2020", subtitle = "Focos de calor por día y acumulados al 10/11/2020, en base a datos VIIRS de FIRMS-NASA", caption = "Natalia Morandeira; 3iA-UNSAM")  

plot_acum


#idem en inglés
plot_cum_eng <- focos_cum %>%
  pivot_longer(names_to = "focos", values_to = "cantidad", col=cantidad_diaria:acumulado) %>%
  ggplot(aes (x=acq_date, y=cantidad, color=focos)) +
  #annotation_custom(rasterGrob(img, width = unit(1,"npc"), height = unit(1,"npc")), 
  #                  -Inf, Inf, -Inf, Inf) +
  geom_line() +
  scale_color_manual(values=c("darkred", "navyblue"),  labels = c("Cummulative hotspots", "New hotspots")) +
  xlab("Date") +
  ylab("Number of VIIRS hotspots") +
  scale_x_date(date_labels = "%m/%d/%Y", breaks = "week") 
  scale_y_continuous(breaks=seq(from = 0, to = (hotspots_count2020 + 1000), by = 2500)) +
  theme_bw()+
  theme(axis.text.x = element_text(angle=45, hjust=1, color="black"), plot.caption = element_text(hjust = 0, vjust=1, face = "italic"), plot.title = element_text(face = "bold"), legend.position=c(0.15, 0.85), legend.title = element_blank(), axis.line=element_line(color="black"), axis.text.y=element_text(color="black") , legend.background = element_rect(colour ="grey40", size = 0.2)) +
  labs(title="Potential fires at the Paraná River Delta (Argentina)", subtitle = "Daily hotspots and cummulative numbers up to 11/10/2020, based on VIIRS data - FIRMS-NASA", caption = "Natalia Morandeira; 3iA-UNSAM")  
plot_cum_eng

```

We may save these plots.
```{r save_cum, eval=F, include=T}
<<<<<<< HEAD
ggsave("output/Focos_acumulados_2020-11-10.png", plot = plot_acum, width = 8, height = 5, dpi = 300)
ggsave("output/Hotspot_cum_2020-11-10.png", plot = plot_cum_eng, width = 8, height = 5, dpi = 300)
=======
ggsave("output/Focos_acumulados_2020-10-04.png", plot = plot_acum, width = 8, height = 5, dpi = 300)
ggsave("output/Hotspot_cum_2020-10-04.png", plot = plot_cum_eng, width = 8, height = 5, dpi = 300)
>>>>>>> a8a7c5ead622c4e9943587b8aabe666ba7da5c88
```

The next plot summarizes the number of VIIRS hotspots per month. 

```{r VIIRS_mes}
focos_VIIRS <- hotspots_all
focos_VIIRS <- clean_names(focos_VIIRS)

focos_VIIRS <- focos_VIIRS %>%
  filter(instrument == "VIIRS") %>%
  mutate(cantidad = 1) %>%
  group_by(acq_date) %>%
  summarize(cantidad_diaria = sum(cantidad)) 

focos_VIIRS2020_mes <- focos_VIIRS %>% 
  filter (acq_date > "2020-01-01") %>%
  mutate(mes = format(acq_date, "%Y/%m")) %>%
  group_by(mes) %>%
  summarize(cantidad_mes = sum(cantidad_diaria))

#convertir a fecha
focos_VIIRS2020_mes$mes <- as.Date(paste(focos_VIIRS2020_mes$mes,1,sep="/"),"%Y/%m/%d") 
head(focos_VIIRS2020_mes) 

VIIRS2020_mes <- ggplot(focos_VIIRS2020_mes, aes(x=mes, y=cantidad_mes)) +
  geom_col( fill="lightblue") +
  geom_text(aes(label = cantidad_mes), col="black") +
  xlab("mes") +
  ylab("Cantidad de focos VIIRS por mes") +
  theme_bw()
VIIRS2020_mes

VIIRSagosto <- (focos_VIIRS2020_mes$cantidad_mes)[8]

ggsave("output/Focos_VIIRS2020_mensual.png", plot = VIIRS2020_mes, width = 8, height = 5, dpi = 300)
```

The month with the highest number of hotspots is August, with **`r VIIRSagosto`** hotspots (**`round(VIIRSagosto/hotspots_count2020*100,1)`%** % of the total hotspots up to **`r last_date`**).


## Historical comparison using MODIS (November 2001 - present)
While VIIRS data (resolution: 375 m) is available from 2012, MODIS data is available from 2001. However, MODIS resolution is 1 km, so fewer number of hotspots are reported and each hotspot corresponds to a greater area. 

```{r MODIS}
focos_MODIS <- hotspots_all

focos_MODIS <- clean_names(focos_MODIS)
colnames(focos_MODIS)

focos_MODIS <- focos_MODIS %>%
  filter(instrument == "MODIS") %>%
  mutate(cantidad = 1) %>%
  group_by(acq_date) %>%
  summarize(cantidad_diaria = sum(cantidad)) 

head(focos_MODIS)
```

Now we compute and plot the number of MODIS hostpots per year.

```{r modis2}
#calculo de focos por anio
focos_MODIS_anio <- focos_MODIS %>% 
  mutate(anio = format(acq_date, "%Y")) %>%
  group_by(anio) %>%
  summarize(cantidad_anio = sum(cantidad_diaria))

glimpse(focos_MODIS_anio)

MODIS_year <- ggplot(subset(focos_MODIS_anio, focos_MODIS_anio$anio!="2000"), aes(x=anio, y=cantidad_anio)) +
  geom_col( fill="lightblue") +
  geom_text(aes(label = cantidad_anio), col="black") +
  xlab("Año") +
  ylab("Cantidad de focos MODIS por año") +
  theme_bw()
MODIS_year

ggsave("output/MODIS_2001-2020.png", plot = MODIS_year, width = 8, height = 5, dpi = 300)

MODIS_2020 <- as.numeric(as.list(subset(focos_MODIS_anio, focos_MODIS_anio$anio ==2020))[2])
MODIS_2008 <-  as.numeric(as.list(subset(focos_MODIS_anio, focos_MODIS_anio$anio ==2008))[2])
```

The cummulative number of MODIS hotspots during 2020 is **`r MODIS_2020`** up to **`r last_date`**, which is **`r round(MODIS_2020/MODIS_2008*100,1)`%** of the MODIS hotspots recorded during 2008. 

Some plots summarizing the historical hotspots from MODIS data. Daily and monthly plots.

```{r MODIS_plots}
#diario
MODIS_diario <- focos_MODIS %>%
  ggplot(aes (x=acq_date, y=cantidad_diaria)) +
  geom_line(color="darkred") +
  xlab("Fecha") +
  ylab("Cantidad de focos de incendio") +
  scale_x_date(date_labels = "%d/%m/%Y", breaks = "year") +
  theme(axis.text.x = element_text(angle=45, hjust=1, color="black"), plot.caption = element_text(hjust = 0, vjust=1, face = "italic"), plot.title = element_text(face = "bold"), legend.position=c(0.12, 0.85), legend.title = element_blank(), axis.line=element_line(color="black"), axis.text.y=element_text(color="black")) +
  labs(title="Datos históricos - incendios en el Delta del Paraná (Argentina)", subtitle = "Focos de calor por día (2001 - presente), en base a datos MODIS de FIRMS-NASA", caption = "Natalia Morandeira; 3iA-UNSAM") 

MODIS_diario

#ggsave("Focos_historicos_diarios.png", plot = p3, width = 8, height = 5, dpi = 300)
```

```{r MODIS_month}
#agrupar dias en meses
focos_MODIS_mes <- focos_MODIS %>% 
  mutate(mes = format(acq_date, "%Y/%m")) %>%
  group_by(mes) %>%
  summarize(cantidad_mes = sum(cantidad_diaria))

#convertir a fecha
focos_MODIS_mes$mes <- as.Date(paste(focos_MODIS_mes$mes,1,sep="/"),"%Y/%m/%d")
head(focos_MODIS_mes)

MODIS_mes <-  focos_MODIS_mes %>%
  ggplot(aes (x=mes, y=cantidad_mes)) +
  geom_line(col="darkred") +
  scale_color_manual(values=c("red", "blue"),  labels = c("Focos acumulados", "Nuevos focos")) +
  xlab("Fecha") +
  ylab("Cantidad de focos de incendio") +
  scale_x_date(date_labels = "%m/%Y", breaks = "6 months") +
  theme(axis.text.x = element_text(angle=45, hjust=1, color="black"), plot.caption = element_text(hjust = 0, vjust=1, face = "italic"), plot.title = element_text(face = "bold"), legend.position=c(0.12, 0.85), legend.title = element_blank(), axis.line=element_line(color="black"), axis.text.y=element_text(color="black")) +
  labs(title="Datos históricos - incendios en el Delta del Paraná (Argentina)", subtitle = "Focos de calor mensuales (2001 - presente), en base a datos MODIS de FIRMS-NASA", caption = "Natalia Morandeira; 3iA-UNSAM") 

MODIS_mes

#ggsave("Focos_historicos_mensual.png", plot = p4, width = 8, height = 5, dpi = 300)
```

Now we can compare VIIRS and MODIS records, from 2012.

```{r VIIRS_all}
focos_VIIRS <- hotspots_all
focos_VIIRS <- clean_names(focos_VIIRS)

focos_VIIRS <- focos_VIIRS %>%
  filter(instrument == "VIIRS") %>%
  mutate(cantidad = 1) %>%
  group_by(acq_date) %>%
  summarize(cantidad_diaria = sum(cantidad)) 

#calculo de focos por anio
focos_VIIRS_anio <- focos_VIIRS %>% 
  mutate(anio = format(acq_date, "%Y")) %>%
  group_by(anio) %>%
  summarize(cantidad_anio = sum(cantidad_diaria))

glimpse(focos_VIIRS_anio)
#write.csv(x = focos_VIIRS_anio, file = "focos_anuales_VIIRS.csv")


VIIRS_anio <- ggplot(subset(focos_VIIRS_anio, focos_VIIRS_anio$anio!="2000"), aes(x=anio, y=cantidad_anio)) +
  geom_col( fill="lightblue") +
  geom_text(aes(label = cantidad_anio), col="black") +
  xlab("Año") +
  ylab("Cantidad de focos VIIRS por año") +
  theme_bw()
VIIRS_anio

ggsave("output/VIIRS_2012-2020.png", plot = VIIRS_anio, width = 8, height = 5, dpi = 300)
```

Lastly, we can comparte VIIRS and MODIS records. Spanish and English versions. 

```{r comparison}
#preparo label del eje X
break.fecha <- c(seq(from = as.Date("2001-01-01"), to = as.Date("2020-11-01"), by = "6 months"), as.Date("2021-01-01"))

#agrupar dias en meses
focos_VIIRS_mes <- focos_VIIRS %>% 
  mutate(mes = format(acq_date, "%Y/%m")) %>%
  group_by(mes) %>%
  summarize(cantidad_mes = sum(cantidad_diaria))

#convertir a fecha
focos_VIIRS_mes$mes <- as.Date(paste(focos_VIIRS_mes$mes,1,sep="/"),"%Y/%m/%d")
head(focos_VIIRS_mes)

MODIS_VIIRS_plot <-  ggplot(data = focos_MODIS_mes, aes (x=mes, y=cantidad_mes)) +
  geom_line(col="darkred") +
  xlab("Fecha") +
  ylab("Cantidad de focos de incendio") +
  scale_x_date(date_labels = "%m/%Y", breaks = break.fecha) +
  theme_bw() +
  theme(axis.text.x = element_text(angle=45, hjust=1, color="black"), plot.caption = element_text(hjust = 0, vjust=1, face = "italic"), plot.title = element_text(face = "bold"), legend.position=c(0.12, 0.85), legend.title = element_blank(), axis.line=element_line(color="black"), axis.text.y=element_text(color="black")) +
  labs(title="Datos históricos - incendios en el Delta del Paraná (Argentina)", subtitle = "Focos de calor mensuales (2001 - presente), en base a datos FIRMS-NASA. Las líneas indican \nfocos detectados por MODIS (1 km) y las barras representan focos VIIRS (375 m)", caption = "Natalia Morandeira; 3iA-UNSAM") 

MODIS_VIIRS_plot<- MODIS_VIIRS_plot + geom_col(data= focos_VIIRS_mes, aes (x=mes, y=cantidad_mes, group=1),  fill="orange", alpha=0.5, width = 30)

MODIS_VIIRS_plot



MODIS_VIIRS_plot_eng <-  ggplot(data = focos_MODIS_mes, aes (x=mes, y=cantidad_mes)) +
  geom_line(col="darkred") +
  xlab("Date") +
  ylab("Number of hotspots") +
  scale_x_date(date_labels = "%m/%Y", breaks = break.fecha) +
  theme_bw() +
  theme(axis.text.x = element_text(angle=45, hjust=1, color="black"), plot.caption = element_text(hjust = 0, vjust=1, face = "italic"), plot.title = element_text(face = "bold"), legend.position=c(0.12, 0.85), legend.title = element_blank(), axis.line=element_line(color="black"), axis.text.y=element_text(color="black")) +
  labs(title="Historical data - Potential fires in the Paraná River Delta (Argentina)", subtitle = "Monthly number of hotspots (2001 - present), based on FIRMS-NASA data. \nLines indicate MODIS hotspots (1 km) and bars indicate VIIRS hotspots (375 m)", caption = "Patricia Kandus, Natalia Morandeira and Priscilla Minotti; 3iA-UNSAM") 

MODIS_VIIRS_plot_eng <- MODIS_VIIRS_plot + geom_col(data= focos_VIIRS_mes, aes (x=mes, y=cantidad_mes, group=1),  fill="orange", alpha=0.5, width = 30)

MODIS_VIIRS_plot_eng


```

We may save these plots.
```{r historical_save, eval=F, include=T}
ggsave("output/Focos_historicos_ambossensores.png", plot = MODIS_VIIRS_plot, width = 8, height = 5, dpi = 300)
ggsave("output/Focos_historicos_ambossensores_english.png", plot = MODIS_VIIRS_plot_eng, width = 8, height = 5, dpi = 300)
```
